\bf A Journey from Wild to Textbook Data to Reproducibly Refresh the Wages Data from the National Longitudinal Survey of Youth Database

### 2022-04-25 paper/paper-jsdse/paper-jsdse.tex
Add all of reviewers' comments in response file.

%\pdfminorversion=4
% NOTE: To produce blinded version, replace "0" with "1" below.
\newcommand{\blind}{1}

@@

Statistics and data science education relies on cleaned and simplified data, suitably called textbook data, for clear examples about how to apply different techniques. An example of this, is the wages data made public by Singer and Willett (2003) in their book, Applied longitudinal data analysis: Modeling change and event occurrence, which can be used to teach generalized linear models, in addition to hierarchical, mixed effects and multilevel models. The data records hourly wages of a sample of high school dropouts, from 1979-1994, along with the demographic variables, such as education and race, taken from the National Longitudinal Survey of Youth (NLSY79) (Bureau of Labor Statistics, U.S. Department of Labor 2021a).

The story from modeling the data (and as reported by Singer and Willett) is that wages increase with the length of time in the workforce, higher level of education leads to higher wages, and that race makes a difference, on average. An exploratory analysis reveals, however, that the individual experience varies a great deal from the overall average. Some individuals experience a decline in wages the longer they are in the workforce, and many experience volatility in their wages. The wages data was used to illustrate exploratory longitudinal data analysis in Ilk (2004), and was further developed into a case study for use in the teaching of exploratory data analysis at CENSORED.

This disparity between the average and the individual is a part of statistics, as a discipline, that require more attention. This particular data set is a prime example of discussing this disparity. Textbook data sets have longevity if there is a unresolved mystery. The iris data (Andereson 1935) is a prime example. It has withstood the test of time because the three species cannot be perfectly classified, and so it continues to challenge researchers and instructors to do better in the analysis. (A side note: the iris data is best replaced today with the penguins data (Horst, Hill, and Gorman 2020), which has similar qualities, is new and does not suffer from a connection with eugenics (Stodel 2020). We argue that the wages data is in this class of textbook data, too, because it presents a challenge for longitudinal data analysis: how can we better summarize and explain the individual experience?

For the field of statistics, and data science by association, it is increasingly important to reach the individual. One might describe this as a divergence of purpose, statistics for public policy or statistics for the public. The two are not the same. As the world becomes more electronically connected combating misinformation and mitigating conspiracy theories require that statistics address the individual. For example, with the wages data, even though the message for public policy is that demographic profile is related to different wage patterns on average, the message for the individual is that you are more than your demographic. The majority of people in the study do not have a pattern that is similar to the average. If you have a bad experience, that your wages have declined over time, you are not alone, there are others like you, and more than you think. A similar tone is echoed occasionally in the public media, for example, an article published in the Sydney Morning Herald argues that there is no average Australian (Moncrief 2015).

As a textbook data set though, the wages data is outdated. The most recent year in the data is 1994, 10 years prior to when Singer and Willett (2003) was published. Teachers of statistics need contemporary data sets to show how techniques are relevant for today's students. Using tired old textbook data sets can imbue a misconception that the field is not current. The wages data is extracted from NSLY79, one of the best examples of open data, which is constantly being updated. It should be possible to continuously refresh the textbook data from the data repository. This paper describes our (non-glamorous) journey from open (Open Knowledge Foundation 2021) wild data to textbook data.

@@

Our process of cleaning builds heavily on the \texttt{tidyverse} approach (Wickham, Averick, et al. 2019). The data is first organized into ``tidy data'' (Wickham 2014) and then further wrangled using the data pipeline and split-apply-combine approach (Wickham 2011). Tidy data shouldn't be confused with ``tame data'' which Kim, Ismay, and Chunn (2018) coined to refer to textbook data sets suitable for teaching, particularly teaching statistics. The resulting (tame) data is provided in a new R package called \texttt{[CENSORED]} which includes the code so that the process is reproducible, and could be used to further refresh the data as new records are made available in the NLSY79 database.

@@

The book ``Applied longitudinal data analysis'' (Singer and Willett 2003) used the wages and other variables of high school dropouts from the NLSY79 data as an example data set to illustrate longitudinal data modeling of wages on workforce experience, with covariates education and race. This data has been playing an important role in research in various disciplines, including but not limited to economics, sociology, education, public policy, and public health for more than a quarter of the century (Pergamit et al. 2001). In addition, this is considered a carefully designed longitudinal survey with high retention rates, making it suitable for life course research (Pergamit et al. 2001; Cooksey 2017). According to Cooksey (2017), thousands of articles, and hundreds of book chapters and monographs have utilized this data. Moreover, the NLSY79 is considered the most widely used and most important cohort in the survey data (Pergamit et al. 2001).
Our aim is to refresh this textbook data and append it with data from 1994 through to the latest data reported in 2018, a purpose that is consistent with Grimshaw (2015)'s statistics education goal of embracing authentic data experiences. Here, we investigate the process of getting from the raw NLSY79 data to a textbook data set as similar as possible to that provided by Singer and Willett (2003). We should also note that race is a variable in the original data set, and for compatibility it is also provided with the refreshed data, for the \emph{purposes of studying racism, not race} (Fullilove 1998).

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  ID: the respondents' ID.
\item
  EXPER: temporal scale, i.e., the length of time (years) in the workforce, starting on the respondents' first day at work.
\item
  LNW: natural logarithm of wages, adjusted with 1990's inflation rate.
\item
  BLACK: binary variable, 1 indicates black and 0 otherwise.
\item
  HISPANIC: binary variable, 1 indicates hispanic and 0 otherwise.
\item
  HGC: the highest grade completed.
\item
  UERATE: unemployment rate of the year of survey. When missing, the variable is set to be 7.875 (the average rate).
\end{enumerate}

We refresh this data by re-creating the full data with records from survey years 1979 through to 2018 (the most recent year published). We also modify some variables. For example, use a single categorical race variable instead of the two binary race variables. We also plan to include additional variables, some for the purpose of providing more options for data exploration in teaching examples: year of the survey, age of individual in 1979, whether the individual completed high school with diploma or with Graduate Equivalency Degree (GED), the number of jobs that the individual had in the corresponding year of survey, total number of hours the individual usually works per week, the year when individual started to work, and the number of years the individual worked. We do not attempt to re-create the unemployment rate variable.

@@

\begin{tcolorbox}[title = Navigating the data source]
\faDatabase\ NLSY79 (\url{https://www.nlsinfo.org/investigator/pages/search?s=NLSY79})\\
\vspace{1mm}
\faCheck\ The CASEID will be always be selected.  \\
\vspace{1mm}
\faCheck\ The 3 recommended demographic variable (sample ID, race and sex) were selected.  \\
\vspace{1mm}
For the remaining variables, we went to the "Variable Search" tab and select variables as follows
\begin{itemize}
\item[$\triangleright$] Education, Training and Achievement Scores
\begin{itemize}
\item[$\triangleright$] Education $\triangleright$ Summary measures $\triangleright$ All schools $\triangleright$ By year
\begin{itemize}
\item[$\triangleright$] Dates of diploma or degree
\begin{itemize}
\item[\faCheck] All variables named Q3-8A were selected. 
\end{itemize}
\end{itemize}
\begin{itemize}
\item[$\triangleright$] Highest grade completed
\begin{itemize}
\item[\faCheck] Selected HGC\_EVER, which is the highest grade respondent has ever reported completed and getting credit for.
\end{itemize}
\end{itemize}
\end{itemize}
\item[$\triangleright$] Employment
\begin{itemize}
\item[$\triangleright$] Summary measures $\triangleright$ By job
\begin{itemize}
\item[$\triangleright$] Hours worked  
\begin{itemize}
\item[\faCheck] All 447 primary variables in Hours worked were selected.
\end{itemize}
\item[$\triangleright$] Hourly wages
\begin{itemize}
\item[\faCheck] All 156 variables in Hourly wages were selected.
\end{itemize}
\end{itemize}
\end{itemize}
\begin{itemize}
\item[$\triangleright$] Summary measures $\triangleright$ Since date of last interview $\triangleright$ Weeks worked
\begin{itemize}
\item[\faCheck] All 28 variables in Weeks worked were selected.
\end{itemize}
\end{itemize}
\begin{itemize}
\item[$\triangleright$] Employer Roster $\triangleright$ Job dates $\triangleright$ Original start date
\begin{itemize}
\item[\faCheck] Only selected the start date (Year) for the first job (E00101.02)
\end{itemize}
\end{itemize}
\item[$\triangleright$] Household, Geography \& Demographics
\begin{itemize}
\item[$\triangleright$] Demographics $\triangleright$ Basic demographics $\triangleright$ Date of birth
\begin{itemize}
\item[\faCheck] All 4 variables in Date of birth were selected. 
\end{itemize}
\end{itemize}
\end{itemize}
\begin{itemize}
\item[\faCloudDownload] To download all 663 variables selected, we then navigate to the tab "Save / Download" then select the tab "Advanced Download". We select the R Source code and Comma-delimited datafile of selected variables with Reference Number as column headers. We name the filename "NLSY79" and press the download button. There are also options to get control or dictionary files for SAS, SPSS and STATA. 
\end{itemize}
\end{tcolorbox}
\caption{Documented steps taken to select variables of interest and download the raw data.\label{fig:source-nav}}
\end{figure}

@@

\begin{itemize}
\tightlist
\item
  \texttt{NLSY79.csv}: Comma Separated Value format of the response data,
\item
  \texttt{NLSY79.dat}: .dat format of the response data,
\item
  \texttt{NLSY79.NLSY79}: Tagset of variables that can be uploaded to the website to recreate the data set, and
\item
  \texttt{NLSY79.R}: R script for reading the data into R and converting the variables' names and label into something more sensible.
\end{itemize}

We alter only the file path in \texttt{NLSY79.R} and run the script without any other alteration. This results in an initial processing of the raw data into two data sets, \texttt{categories\_qnames} (where the observations are stored in categorical/interval values) and \texttt{new\_data\_qnames} (the observations are stored in integer form).

According to Wickham (2014), tidy data sets comply with three rules: (i) each variable forms a column, (ii) each observation forms a row, and (iii) each type of observational unit forms a table. The raw data, \texttt{new\_data\_qnames}, does not comply with these rules as it is organised such that each row corresponds to an individual. As respondents can have multiple jobs at specific years, the column names, such as \texttt{HRP1\_1979}, \texttt{HRP2\_1979}, \texttt{HRP1\_1980} and \texttt{HRP2\_1980}, contain the information about the job number up to 5 (\texttt{HRP1} = job 1, \texttt{HRP2} = job 2) and the year. The raw data consequently has a large number of columns (742 to be specific). The values in the cell under the variables that begin with \texttt{HRP} correspond to the hourly wage in dollars. A glimpse of this data shows:

\begin{verbatim}
#> 'data.frame':    12686 obs. of  742 variables:
#>  $ CASEID_1979                               : int  1 2 3 4 5 6 7 8 ...
#>  $ HRP1_1979                                 : int  328 385 365 NA 310 NA NA NA ...
#>  $ HRP2_1979                                 : int  NA NA NA NA 375 NA NA NA ...
#>  $ HRP3_1979                                 : int  NA NA 275 NA NA NA NA NA ...
#>  $ HRP4_1979                                 : int  NA NA NA NA NA 250 NA NA ...
#>  $ HRP5_1979                                 : int  NA NA NA NA NA NA NA NA ...
#>  $ HRP1_1980                                 : int  NA 457 397 NA 333 275 300 394 ...
#>  $ HRP2_1980                                 : int  NA NA 367 NA NA NA NA NA ...
#>  $ HRP3_1980                                 : int  NA NA 380 NA NA NA 290 NA ...
#>  $ HRP4_1980                                 : int  NA NA NA NA NA NA NA NA ...
#>   [list output truncated]
\end{verbatim}

Thus we re-arrange and wrangle the data into tidy data form, columns corresponding to individual ID, year, job number, wage in dollars and the demographic variables. This is done using the \texttt{tidyverse} suite of packages (Wickham, Averick, et al. 2019): \texttt{tidyr} (Wickham 2020) to pivot the data into long form, with \texttt{dplyr} (Wickham, François, et al. 2020) and \texttt{stringr} (Wickham 2019) to mutate new variables from the downloaded data from the database, and code levels of factors by text wrangling. The long form of the data makes it possible to do these data transformations efficiently, and it is an intermediate step towards the final target data. The code for tidying the data are demonstrated at \url{CENSORED/articles/raw-to-input-data.html} but also described in the subsequent subsections.

@@

In our final target data, we wish to include the demographic variables with variable names specified in brackets: gender (\texttt{gender}), race (\texttt{race}), age (\texttt{age\_1979}), highest grade completed (\texttt{hgc}), highest grade completed in terms of years, e.g.~9th grade = 9, 3rd year college = 15, (\texttt{hgc\_i}), and whether the graduate equivalency diploma is obtained (\texttt{ged}).

For gender and race, we only rename the column names. It is worth noting that using these two variable needs special attention. Gender, as reported in the data, only has two categories, which is recognised today as inadequate. Gender is not binary. Further, race is as reported in the database. When doing analysis with this variable, one should keep in mind that the purpose is to study racism rather than race.

The raw data, \texttt{new\_data\_qnames}, contains the variables \texttt{Q1-3\_A\textasciitilde{}Y\_1979} and \texttt{Q1-3\_A\textasciitilde{}Y\_1981} which records two versions of the birth year of the respondent; this is also the case for the record of birth month (\texttt{Q1-3\_A\textasciitilde{}M\_1979} and \texttt{Q1-3\_A\textasciitilde{}M\_1981}). The record contains two versions of birth year and birth month as the survey recorded this in 1979 and 1981. We checked for consistency between the two versions and found no discrepancy where the responses were recorded in both 1979 and 1981. The age was then calculated using the birth year.

The next step is tidying to obtain \texttt{hgc} and \texttt{ged}. Besides that \texttt{hgc} is available in the original data, along with \texttt{ged}, it is used to subset the data to the high school dropouts as in the original data. As the highest grade completed is already exist in \texttt{new\_data\_qnames} and stored in year unit (e.g., 10, 11, 12, 13, and so on), we only rename the column name to be \texttt{hgc\_i} and recode it so we have the factor data type (e.g., 10th grade, 11th grade, and so on). For \texttt{ged}, we only separate the year of survey and the GED status of the individual.

@@

For \texttt{stwork} variable, we only rename the column names from \texttt{new\_data\_qnames}. This variable is then used to calculate the next variable, \texttt{yr\_wforce} for each survey round, which is the year of survey (\texttt{year}) minus the year of individual start working (\texttt{stwork}). Finally, \texttt{exp} variable is derived from the number of weeks worked since the last interview indicated as variable started with \texttt{WKSWK} in \texttt{new\_data\_qnames}. To obtain the work experience since 1979, we calculate the cumulative value. As the measurement unit is in week, we convert this to year.

The employment and demographic variables are then joined. These data are further filtered to the cohort who participated in at least three rounds in the survey, the minimum observation recommended for longitudinal data in Singer and Willett (2003). We also opt to restrict the minimum number of observations so that the data can be used to demonstrate within-person variation when it is used to teach longitudinal data. However, it is worth noting that the original data does not restrict the number of observation for each individual, i.e, there are individuals with only one and two observations.

Finally, we save the resultant wage data on this cohort as \texttt{wages}.

@@

Work experience is one of the most important variables in Singer and Willett (2003) as it indicates time, and makes longitudinal analysis possible. It is desirable to calculate this rather than using the survey year for time because it more accurately reflects a person's time in the workforce. Thus, in the spirit of refreshing the data to the newest round of the survey, this variable needs to be calculated from other variables provided. It is not straightforward. We start with the definition of experience in Singer and Willett (2003).

Experience is years after entering the labor force. It represents the difference between the day an individual enters the labor force (\texttt{EMPLOYERS\_ALL\_STARTDATE\_ORIGINAL.01\textasciitilde{}Y\_XRND} in the database) relative to the date of the survey, which we call \texttt{yr\_wforce}. However, using this calculation produced numbers which don't quite match with the original data.

Reading the section titled \href{https://www.nlsinfo.org/content/cohorts/nlsy79/topical-guide/employment/work-experience}{``Topical Guide to the Data''} in the guide, suggests that it should be calculated based on the variable ``number of weeks worked since last interview''. This would remove periods of unemployment which makes sense when measuring experience while actually working. Since it is only measured since the last interview, this needs to be cumulated for each survey year. This produced results more similar to the original data (see Subsection \ref{takeaways}).

@@

\caption{Longitudinal profiles of wages for a random sample of 20 individuals in the refreshed data. Most, but not all, individuals here experienced increasing wages over time, and several have experienced considerable fluctuation in wages. Some individuals are only measured for a short period.}\label{fig:sample-plot}
\end{figure}

>>>>

%\pdfminorversion=4
% NOTE: To produce blinded version, replace "0" with "1" below.
\newcommand{\blind}{0}

@@

Statistics and data science education relies on cleaned and simplified data, suitably called textbook data, for clear examples about how to apply different techniques. An example of this is the wages data made public by Singer and Willett (2003) in their book, ``Applied longitudinal data analysis'', which can be used to teach generalized linear models, in addition to hierarchical, mixed effects, and multilevel models. The data records hourly wages of a sample of high school dropouts, from 1979-1994, along with the demographic variables, such as education and race, taken from the National Longitudinal Survey of Youth (NLSY79) (Bureau of Labor Statistics, U.S. Department of Labor 2021a).

The story from modeling the data (and as reported by Singer and Willett) is that wages increase with the length of time in the workforce, a higher level of education leads to higher wages, and that race makes a difference, on average. An exploratory analysis reveals, however, that the individual experience varies a great deal from the overall average. Some individuals experience a decline in wages the longer they are in the workforce and many experience volatility in their wages. The wages data was used to illustrate exploratory longitudinal data analysis in Ilk (2004), and was further developed into a case study for use in the teaching of exploratory data analysis at CENSORED.

This disparity between the average and the individual is a part of statistics, as a discipline that requires more attention. This particular data set is a prime example of discussing this disparity. Textbook data sets have longevity if there is an unresolved mystery. The iris data (Andereson 1935) is a prime example. It has withstood the test of time because the three species cannot be perfectly classified, and so it continues to challenge researchers and instructors to do better in the analysis. (A side note: the iris data is best replaced today with the penguins data (Horst, Hill, and Gorman 2020), which has similar qualities, is new, and does not suffer from a connection with eugenics (Stodel 2020). We argue that the wages data is in this class of textbook data, too, because it presents a challenge for longitudinal data analysis: how can we better summarize and explain the individual experience?

For the field of statistics and data science by association, it is increasingly important to reach the individual. One might describe this as a divergence of purpose, statistics for public policy, or statistics for the public. The two are not the same. As the world becomes more electronically connected, combating misinformation and mitigating conspiracy theories require that statistics address the individual. For example, with the wages data, even though the message for public policy is that demographic profile is related to different wage patterns on average, the message for the individual is that you are more than your demographic. The majority of people in the study do not have a pattern that is similar to the average. If you have a bad experience, that your wages have declined over time, you are not alone; there are others like you, and more than you think. A similar tone is echoed occasionally in the public media. For example, an article published in the Sydney Morning Herald argues that there is no average Australian (Moncrief 2015).

As a textbook data set, though, the wages data is outdated. The most recent year in the data is 1994, 10 years prior to when Singer and Willett (2003) was published. Teachers of statistics need contemporary data sets to show how techniques are relevant for today's students. Using tired old textbook data sets can imbue a misconception that the field is not current. The wages data is extracted from NSLY79, one of the best examples of open data, which is constantly being updated. It should be possible to continuously refresh the textbook data from the data repository. This paper describes our (non-glamorous) journey from open (Open Knowledge Foundation 2021) wild data to textbook data.

@@

Our approach to cleaning builds heavily on the \texttt{tidyverse} approach (Wickham, Averick, et al. 2019). The data is first organized into ``tidy data'' (Wickham 2014) and then further wrangled using step-wise piping with a split-apply-combine strategy for mutating new variables (Wickham 2011). Tidy data shouldn't be confused with ``tame data'', which Kim, Ismay, and Chunn (2018) coined to refer to textbook data sets suitable for teaching, particularly teaching statistics. The resulting (tame) data is provided in a new R package called \texttt{[CENSORED]}, which includes the code so that the process is reproducible and could be used to further refresh the data as new records are made available in the NLSY79 database.

@@

Singer and Willett (2003) used the wages and other variables of high school dropouts from the NLSY79 data as an example data set to illustrate longitudinal data modeling of wages on workforce experience, with covariates education and race. This data has been playing an important role in research in various disciplines, including but not limited to economics, sociology, education, public policy, and public health for more than a quarter of the century (Pergamit et al. 2001). In addition, this is considered a carefully designed longitudinal survey with high retention rates, making it suitable for life course research (Pergamit et al. 2001; Cooksey 2017). According to Cooksey (2017), thousands of articles and hundreds of book chapters and monographs have utilized this data. Moreover, the NLSY79 is considered the most widely used and most important cohort in the survey data (Pergamit et al. 2001).

Our aim is to refresh the wages textbook data and append it with data from 1994 through to the latest data reported in 2018, a purpose that is consistent with Grimshaw (2015)'s statistics education goal of embracing authentic data experiences. Here, we investigate the process of getting from the raw NLSY79 data to a textbook data set as similar as possible to that provided by Singer and Willett (2003). We should also note that race is a variable in the original data set, and for compatibility, it is also provided with the refreshed data for the \emph{purposes of studying racism, not race} (Fullilove 1998). There are a number of data sets provided by Singer and Willett (2003), and we focus only on this one because it has captivated our attention for a number of years. We use it in our own teaching of longitudinal data analysis and would very much like to use data since 1994.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  ID: the respondents' ID.
\item
  EXPER: temporal scale, i.e., the length of time (years) in the workforce, starting on the respondents' first day at work.
\item
  LNW: natural logarithm of wages, adjusted with 1990's inflation rate.
\item
  BLACK: binary variable, 1 indicates black and 0 otherwise.
\item
  HISPANIC: binary variable, 1 indicates hispanic and 0 otherwise.
\item
  HGC: the highest grade completed.
\item
  UERATE: the unemployment rate of the year of the survey. When missing, the variable is set to be 7.875 (the average rate).
\end{enumerate}

We refresh this data by re-creating the full data with records from survey years 1979 through to 2018 (the most recent year published). We also modify some variables. For example, use a single categorical race variable instead of the two binary race variables. We also plan to include additional variables, some for the purpose of providing more options for data exploration in teaching examples: year of the survey, age of individual in 1979, whether the individual completed high school with a diploma or with a graduate equivalency degree (GED), the highest grade completed in the corresponding year of survey, the number of jobs that the individual had in the corresponding year of survey, the total number of hours the individual usually works per week, the year when the individual started to work, and the number of years the individual worked. We do not attempt to re-create the unemployment rate variable.

@@

\begin{tcolorbox}[title = Navigating the data source]
\faDatabase\ NLSY79 (\url{https://www.nlsinfo.org/investigator/pages/search?s=NLSY79})\\
\vspace{1mm}
\faCheck\ The CASEID will be always be selected.  \\
\vspace{1mm}
\faCheck\ The 3 recommended demographic variable (sample ID, race and sex) were selected.  \\
\vspace{1mm}
For the remaining variables, we went to the "Variable Search" tab and select variables as follows
\begin{itemize}
\item[$\triangleright$] Education, Training and Achievement Scores
\begin{itemize}
\item[$\triangleright$] Education $\triangleright$ Summary measures $\triangleright$ All schools $\triangleright$ By year
\begin{itemize}
\item[$\triangleright$] Highest grade completed
\begin{itemize}
\item[\faCheck] All 80 variables in Highest grade completed were selected.
\end{itemize}
\end{itemize}
\begin{itemize}
\item[$\triangleright$] Dates of diploma or degree
\begin{itemize}
\item[\faCheck] All variables named Q3-8A were selected.
\end{itemize}
\end{itemize}
\end{itemize}
\item[$\triangleright$] Employment
\begin{itemize}
\item[$\triangleright$] Summary measures $\triangleright$ By job
\begin{itemize}
\item[$\triangleright$] Hours worked
\begin{itemize}
\item[\faCheck] All 447 primary variables in Hours worked were selected.
\end{itemize}
\item[$\triangleright$] Hourly wages
\begin{itemize}
\item[\faCheck] All 156 variables in Hourly wages were selected.
\end{itemize}
\end{itemize}
\end{itemize}
\begin{itemize}
\item[$\triangleright$] Summary measures $\triangleright$ Since date of last interview $\triangleright$ Weeks worked
\begin{itemize}
\item[\faCheck] All 28 variables in Weeks worked were selected.
\end{itemize}
\end{itemize}
\begin{itemize}
\item[$\triangleright$] Employer Roster $\triangleright$ Job dates $\triangleright$ Original start date
\begin{itemize}
\item[\faCheck] Only selected the start date (Year) for the first job (E00101.02)
\end{itemize}
\end{itemize}
\item[$\triangleright$] Household, Geography \& Demographics
\begin{itemize}
\item[$\triangleright$] Demographics $\triangleright$ Basic demographics $\triangleright$ Date of birth
\begin{itemize}
\item[\faCheck] All 4 variables in Date of birth were selected.
\end{itemize}
\end{itemize}
\end{itemize}
\begin{itemize}
\item[\faCloudDownload] To download all 742 variables selected, we then navigate to the tab "Save / Download" then select the tab "Advanced Download". We select the R Source code and comma-delimited datafile of selected variables with Reference Number as column headers. We name the filename "NLSY79" and press the download button. There are also options to get control or dictionary files for SAS, SPSS and STATA.
\end{itemize}
\end{tcolorbox}
\caption{Documented steps taken to select variables of interest and download the raw data.\label{fig:source-nav}}
\end{figure}

@@

\begin{itemize}
\tightlist
\item
  \texttt{NLSY79.csv}: comma separated value format of the response data,
\item
  \texttt{NLSY79.dat}: alternative text format of the response data,
\item
  \texttt{NLSY79.NLSY79}: tagset of variables that can be uploaded to the website to recreate the data set, and
\item
  \texttt{NLSY79.R}: R script for reading the data into R and converting the variables' names and label into something more sensible.
\end{itemize}

We alter only the file path in \texttt{NLSY79.R} and run the script without any other alteration. This results in the initial processing of the raw data into two data sets, \texttt{categories\_qnames} (where the observations are stored in categorical/interval values) and \texttt{new\_data\_qnames} (the observations are stored in integer form).

According to Wickham (2014), tidy data sets comply with three rules: (i) each variable forms a column, (ii) each observation forms a row, and (iii) each type of observational unit forms a table. The raw data, \texttt{new\_data\_qnames}, does not comply with these rules as it is organized such that each row corresponds to an individual. As respondents can have multiple jobs at specific years, the column names, such as \texttt{HRP1\_1979}, \texttt{HRP2\_1979}, \texttt{HRP1\_1980} and \texttt{HRP2\_1980}, contain the information about the job number up to 5 (\texttt{HRP1} = job 1, \texttt{HRP2} = job 2) and the year. The raw data consequently has a large number of columns (770 to be specific). The values in the cell under the variables that begin with \texttt{HRP} correspond to the hourly wage in dollars. A glimpse of this data shows:

\begin{verbatim}
#> 'data.frame':    12686 obs. of  770 variables:
#>  $ CASEID_1979                               : int  1 2 3 4 5 6 7 8 ...
#>  $ HRP1_1979                                 : int  328 385 365 NA 310 NA NA NA ...
#>  $ HRP2_1979                                 : int  NA NA NA NA 375 NA NA NA ...
#>  $ HRP3_1979                                 : int  NA NA 275 NA NA NA NA NA ...
#>  $ HRP4_1979                                 : int  NA NA NA NA NA 250 NA NA ...
#>  $ HRP5_1979                                 : int  NA NA NA NA NA NA NA NA ...
#>  $ HRP1_1980                                 : int  NA 457 397 NA 333 275 300 394 ...
#>  $ HRP2_1980                                 : int  NA NA 367 NA NA NA NA NA ...
#>  $ HRP3_1980                                 : int  NA NA 380 NA NA NA 290 NA ...
#>  $ HRP4_1980                                 : int  NA NA NA NA NA NA NA NA ...
#>   [list output truncated]
\end{verbatim}

Thus, we re-arrange and wrangle the data into tidy data form, columns corresponding to individual ID, year, job number, wage in dollars, and the demographic variables. This is done using the \texttt{tidyverse} suite of packages (Wickham, Averick, et al. 2019): \texttt{tidyr} (Wickham 2020) to pivot the data into long-form, with \texttt{dplyr} (Wickham, François, et al. 2020) and \texttt{stringr} (Wickham 2019) to mutate new variables from the downloaded data from the database, and code levels of factors by text wrangling. The long form of the data makes it possible to do these data transformations efficiently, and it is an intermediate step towards the final target data. The code for tidying the data are demonstrated at \url{CENSORED/articles/raw-to-input-data.html} but also described in the subsequent subsections.

@@

In our final target data, we wish to include the demographic variables with variable names specified in brackets: gender (\texttt{gender}), race (\texttt{race}), age (\texttt{age\_1979}), highest grade completed reported in each round of the survey (\texttt{grade}), highest grade completed ever reported (\texttt{hgc}), highest grade completed in terms of years, e.g., 9th grade = 9, 3rd-year college = 15, (\texttt{hgc\_i}), highest grade completed in 1979 (\texttt{hgc\_1979}) and whether the graduate equivalency diploma is obtained (\texttt{ged}).

For \texttt{gender} and \texttt{race}, we only rename the column names provided in the raw data, \texttt{new\_data\_qnames}. It is worth noting that \texttt{gender} comes from a variable called \texttt{sex} in the database. We use the term \texttt{gender} as the reports about the data on various pages in the website more commonly use \texttt{gender}. Gender, as provided in the data, is self-reported and only has two categories, ``male'' and ``female''. The interchangeable use of these terms reflects that gender and sex are sometimes regarded as similar in surveys, although they are not. Measuring gender as a binary variable has the potential to fail to capture people who do not identify themselves as either male or female or people whose gender does not align with their sex classification (Kennedy et al. 2020). From a statistical perspective, this sometimes difficult the adjustment of survey statistics to the population when gender is measured, for example, with three categories, and the census measured it with only two categories and used the term sex (Kennedy et al. 2020). Hence, for modern societal studies, it is highly recommended to separate sex and gender in the survey question and provide more options to respond to questions on gender. This discourse should also be emphasized in teaching that using these two variables should be accompanied by special attention. Similarly, this also relates to race, as reported in the database. When doing an analysis with this variable, one should keep in mind that the purpose is to study racism rather than race.

The object \texttt{new\_data\_qnames}, contains the variables \texttt{Q1-3\_A\textasciitilde{}Y\_1979} and \texttt{Q1-3\_A\textasciitilde{}Y\_1981}, which records two versions of the birth year of the respondent; this is also the case for the record of birth month (\texttt{Q1-3\_A\textasciitilde{}M\_1979} and \texttt{Q1-3\_A\textasciitilde{}M\_1981}). The record contains two versions of birth year and birth month as the survey recorded this in 1979 and 1981. We checked for consistency between the two versions and found no discrepancy where the responses were recorded in both 1979 and 1981. The age was then calculated using the birth year.

The next step is processing the highest grade completed. There are several ways to define this, and this should be reflected in the refreshed data to give some flexibility for downstream analysis. The first one is the highest grade ever completed is reported in the database and provided in the refreshed data as \texttt{hgc} and \texttt{hgc\_i}, for the factor and integer type, respectively. For each individual, there is only one value of \texttt{hgc} and \texttt{hgc\_i}. This variable is obtained from \texttt{new\_data\_qnames} with the name \texttt{HGC\_EVER\_XRND} and stored in year units (e.g., 10, 11, 12, 13, and so on), so the transformation is simply giving a new column name as \texttt{hgc\_i} and recoded as a factor to give \texttt{hgc} (e.g., 10th grade, 11th grade, and so on).

The second definition is the highest grade completed that is reported in each round of the survey (provided in the refreshed data as \texttt{grade}). This value can change over time, but it should only increase. This has been included in the refreshed dataset to enable richer downstream analyses, for example, if one would like to explore how the temporal changes in education level affect the wages of individuals. This is recorded in \texttt{new\_data\_qnames} as columns beginning with \texttt{Q3-4} and \texttt{HGC} with year as a suffix. In addition, it is also in the columns beginning with \texttt{HGCREV} reflecting revised data. We chose to use these revised values because there were fewer missing values indicating it had been more thoroughly checked. When this value was missing, 2012, 2014, 2016, and 2018, but available in the first form substituted accordingly.

The third definition is the highest grade completed in 1979 (provided in the refreshed data as \texttt{hgc\_1979}), corresponding to the value from the first round of the survey. This is calculated from the \texttt{grade} value in 1979. This best reflects the grade when the individual left school and is included in order to compare with the original data.

The next step is tidying to obtain \texttt{ged}. Along with \texttt{hgc}, \texttt{ged} is used to subset the data to the high school dropouts as in the original data. The graduate equivalency status is saved as a variable started with ``Q3-8A'' followed by the year of the survey. Thus, we only separate the year and the GED status. Although the GED status is asked in each round of the survey, we only retain the latest status of one's GED.

@@

For \texttt{stwork} variable, we only rename the column names from \texttt{new\_data\_qnames}. This variable is then used to calculate the next variable, \texttt{yr\_wforce} for each survey round, which is the year of survey (\texttt{year}) minus the year of individual started working (\texttt{stwork}). Finally, \texttt{exp} variable is derived from the number of weeks worked since the last interview indicated as variable started with \texttt{WKSWK} in \texttt{new\_data\_qnames}. To obtain the work experience since 1979, we calculate the cumulative value. As the measurement unit is in week, we convert this to year.

The employment and demographic variables are then joined. These data are further filtered to the cohort who participated in at least three rounds in the survey, the minimum observation recommended for longitudinal data in Singer and Willett (2003). We also opt to restrict the minimum number of observations so that the data can be used to demonstrate within-person variation when it is used to teach longitudinal data. However, it is worth noting that the original data does not restrict the number of observations for each individual, i.e., there are individuals with only one and two observations.

Finally, we save the resultant wage data on this cohort as \texttt{wages}. Note that we save \texttt{grade} variable in this dataset instead of \texttt{demog\_nlsy79} dataset because it is a longitudinal variable, while \texttt{demog\_nlsy79} is a cross-sectional dataset reflecting the state of individuals corresponding to the most recent round of the survey.

@@

Work experience is one of the most important variables in Singer and Willett (2003) as it indicates time and makes longitudinal analysis possible. It is desirable to calculate this rather than using the survey year for time because it more accurately reflects a person's time in the workforce. Thus, in the spirit of refreshing the data to the newest round of the survey, this variable needs to be calculated from other variables provided. It is not straightforward. We start with the definition of experience in Singer and Willett (2003).

Experience is years after entering the labor force. It represents the difference between the day an individual enters the labor force (\texttt{EMPLOYERS\_ALL\_STARTDATE\_ORIGINAL.01\textasciitilde{}Y\_XRND} in the database) relative to the date of the survey, which we call \texttt{yr\_wforce}. However, using this calculation produced numbers that do not quite match with the original data.

Reading the section titled \href{https://www.nlsinfo.org/content/cohorts/nlsy79/topical-guide/employment/work-experience}{``Topical Guide to the Data''} in the guide suggests that it should be calculated based on the variable ``number of weeks worked since the last interview''. This would remove periods of unemployment which makes sense when measuring experience while actually working. Since it is only measured since the last interview, this needs to be cumulated for each survey year. This produced results more similar to the original data (as discussed further in Section \ref{takeaways}).

@@

\caption{Longitudinal profiles of wages for a random sample of 36 individuals in the pre-cleaned data. There is considerable variation in wages. Some individuals  (2799, 11041, 11146) are only measured for a short period. Some individuals (8296, 9962) possibly have errors in wages in some years, because of the extreme fluctuation.}\label{fig:sample-plot}
\end{figure}
